---
title: "Epiworld Presentation"
author: "Derek Meyer"
date: "October, 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Overview 
Epiworld is an Agent Based coding package developed with the sole purpose of simulating infectious diseases 
as they spread through populations. The user is able to specify certain parameters in order to
replicate real-world conditions such as initial population size, vaccination rates, hospitalization 
rates, schools as entities, and of course, the disease of interest. Epiworld contains several 
epidemiological models in which the user may select to run such a simulation. These models include:
SIR, SIRConnected, SIS, SEIR, SEIRConnected, and Surveillance.  
  
Epiworld makes use of agent based modeling. That is, it is a computational model 
with the purpose of simulating the interactions of autonomous agents (people). Due to the ABM structure
of epiworld, the ability to analyze the effects of diseases like COVID and the Flu on populations given 
specified parameters reflecting human interaction can be done in seconds. 

##Simulation
#(COVID/FLU in SLC - Parameters: schools as entities, population = 500k, Family distribution in Damon's paper, Vaccination rates, SEIR model+hospitalization)
```{r}

```

