---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| label: setup
#| echo: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# measles

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://github.com/UofUEpiBio/measles/blob/master/LICENSE.md)
<!-- badges: end -->

## Overview

The **measles** package is a specialized spinoff from [epiworldR](https://github.com/UofUEpiBio/epiworldR), focusing exclusively on measles epidemiological models. This package provides fast, agent-based models (ABMs) for studying measles transmission dynamics, vaccination strategies, and intervention policies.

Built on the powerful [epiworld](https://github.com/UofUEpiBio/epiworld) C++ library, these models leverage the speed and flexibility of epiworldR while providing specialized functionality for measles outbreak modeling.

## Features

- **Fast simulation**: Leverages the high-performance C++ backend from epiworld
- **Specialized measles models**: Three distinct models tailored for different scenarios
- **Flexible interventions**: Support for vaccination, quarantine, isolation, and contact tracing
- **Population mixing**: Models can account for different population groups with varying contact patterns
- **Risk-based strategies**: Advanced models support risk-stratified quarantine policies

## Models Included

The package includes three measles-specific models:

1. **ModelMeaslesSchool**: A SEIHR (Susceptible-Exposed-Infectious-Hospitalized-Recovered) model designed for school settings with isolation and quarantine policies.

2. **ModelMeaslesMixing**: A measles model with population mixing between different groups, including vaccination, quarantine, isolation, and contact tracing mechanisms.

3. **ModelMeaslesMixingRiskQuarantine**: An advanced mixing model with risk-based quarantine strategies that assign different quarantine durations based on exposure risk levels (high, medium, low).

## Installation

You can install the measles package from GitHub:

```r
# install.packages("devtools")
devtools::install_github("UofUEpiBio/measles")
```

Note: This package requires [epiworldR](https://cran.r-project.org/package=epiworldR) as a dependency.

## Quick Example

Here's a simple example using the ModelMeaslesSchool:

```{r example, eval=FALSE}
library(measles)

# Create a measles model for a school with 500 students
model_school <- ModelMeaslesSchool(
  n = 500,
  prevalence = 1,
  prop_vaccinated = 0.70,
  contact_rate = 15,
  transmission_rate = 0.9
)

# Run the simulation
run(model_school, ndays = 100, seed = 1912)

# View results
summary(model_school)
plot(model_school)
```

## Example with Population Mixing

The mixing models allow you to simulate measles spread across different population groups:

```{r mixing-example, eval=FALSE}
library(measles)

# Define three populations
e1 <- entity("Population 1", 3000, as_proportion = FALSE)
e2 <- entity("Population 2", 3000, as_proportion = FALSE)
e3 <- entity("Population 3", 3000, as_proportion = FALSE)

# Define contact matrix (row-stochastic: rows sum to 1)
contact_matrix <- matrix(c(
  0.9, 0.05, 0.05,
  0.1, 0.8,  0.1,
  0.1, 0.2,  0.7
), byrow = TRUE, nrow = 3)

# Create the model
measles_model <- ModelMeaslesMixing(
  n = 9000,
  prevalence = 1 / 9000,
  contact_rate = 15,
  transmission_rate = 0.9,
  vax_efficacy = 0.97,
  prop_vaccinated = 0.95,
  contact_matrix = contact_matrix,
  quarantine_period = 14,
  isolation_period = 10
)

# Add entities and run
measles_model |>
  add_entity(e1) |>
  add_entity(e2) |>
  add_entity(e3)

run(measles_model, ndays = 100, seed = 331)
summary(measles_model)
```

## Relationship to epiworldR

This package is a spinoff from epiworldR, which provides a comprehensive framework for agent-based epidemiological models. While epiworldR includes many different disease models (SIR, SEIR, SIS, etc.), the measles package focuses specifically on measles-related models with specialized features for:

- Measles-specific disease progression (incubation, prodromal, and rash periods)
- School-based outbreak scenarios
- Vaccination coverage and efficacy
- Quarantine and isolation policies
- Contact tracing strategies
- Risk-stratified interventions

For general-purpose epidemiological modeling or other disease types, please see the [epiworldR package](https://github.com/UofUEpiBio/epiworldR).

## Citation

If you use the measles package in your research, please cite both this package and epiworldR:

```r
citation("measles")
citation("epiworldR")
```

## Authors

This package was developed as part of a collaboration between the University of Utah (ForeSITE center grant) and the Utah Department of Health and Human Services in response to the 2025 US Measles outbreak.

## License

MIT License - see LICENSE file for details.

## Contributing

Contributions are welcome! Please see the [epiworldR development guidelines](https://github.com/UofUEpiBio/epiworldR/blob/main/DEVELOPMENT.md) for information on how to contribute.
