---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# epiworldR

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/epiworldR)](https://CRAN.R-project.org/package=epiworldR)
[![R-CMD-check](https://github.com/UofUEpiBio/epiworldR/actions/workflows/r.yml/badge.svg)](https://github.com/UofUEpiBio/epiworldR/actions/workflows/r.yml)
[![CRANlogs downloads](https://cranlogs.r-pkg.org/badges/grand-total/epiworldR)](https://cran.r-project.org/package=epiworldR)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://github.com/UofUEpiBio/epiworldR/blob/master/LICENSE.md)
[![codecov](https://codecov.io/gh/UofUEpiBio/epiworldR/graph/badge.svg?token=ZB8FVLI7GN)](https://app.codecov.io/gh/UofUEpiBio/epiworldR)
<!-- badges: end -->

This R package is a wrapper of the C++ library [epiworld](https://github.com/UofUEpiBio/epiworld){target="_blank"}. It provides a general framework for modeling disease transmission using [agent-based models](https://en.wikipedia.org/w/index.php?title=Agent-based_model&oldid=1153634802){target="_blank"}. Some of the main features include:


- Fast simulation with an average of 30 million agents/day per second.
- One model can include multiple diseases.
- Policies (tools) can be multiple and user-defined.
- Transmission can be a function of agents' features.
- Out-of-the-box parallelization for multiple simulations.

From the package's description:

> A flexible framework for Agent-Based Models (ABM), the epiworldR package provides methods for prototyping disease outbreaks and transmission models using a C++ backend, making it very fast. It supports multiple epidemiological models, including the Susceptible-Infected-Susceptible (SIS), Susceptible-Infected-Removed (SIR), Susceptible-Exposed-Infected-Removed (SEIR), and others, involving arbitrary mitigation policies and multiple-disease models. Users can specify infectiousness/susceptibility rates as a function of agents’ features, providing great complexity for the model dynamics. Furthermore, epiworldR is ideal for simulation studies featuring large populations.

Current available models:

```{r print-models, echo=FALSE, results='asis'}
models <- list.files(path = "R/", pattern = "Model.*.R", full.names = FALSE) |>
  gsub(pattern = "(Model.*)\\.R", replacement = "\\1")

sprintf("%i. `%s`\n", 1:length(models), models) |>
  cat()
```

## Installation

You can install the development version of epiworldR from [GitHub](https://github.com/) with:

``` r
devtools::install_github("UofUEpiBio/epiworldR")
```

Or from CRAN

```r
install.packages("epiworldR")
```

# Examples

This R package includes several popular epidemiological models, including
[SIS](https://en.wikipedia.org/w/index.php?title=Compartmental_models_in_epidemiology&oldid=1155757336#Variations_on_the_basic_SIR_model){target="_blank"},
[SIR](https://en.wikipedia.org/w/index.php?title=Compartmental_models_in_epidemiology&oldid=1155757336#The_SIR_model){target="_blank"}, and
[SEIR](https://en.wikipedia.org/w/index.php?title=Compartmental_models_in_epidemiology&oldid=1155757336#The_SEIR_model){target="_blank"} using either a fully connected graph (similar to a compartmental model) or a user-defined network.

## SIR model using a random graph

This Susceptible-Infected-Recovered model features a population of 100,000 agents simulated in a small-world network. Each agent is connected to ten other agents. One percent of the population has the virus, with a 70% chance of transmission. Infected individuals recover at a 0.3 rate:

```{r example}
library(epiworldR)

# Creating a SIR model
sir <- ModelSIR(
  name              = "COVID-19",
  prevalence        = .01,
  transmission_rate = .7,
  recovery          = .3
) |>
  # Adding a Small world population
  agents_smallworld(n = 100000, k = 10, d = FALSE, p = .01) |>
  # Running the model for 50 days
  run(ndays = 50, seed = 1912)

sir
```

Visualizing the outputs

```{r sir-figures}
summary(sir)
plot(sir)
plot_incidence(sir)
```

## SEIR model with a fully connected graph

The SEIR model is similar to the SIR model but includes an exposed state. Here, we simulate a population of 10,000 agents with a 0.01 prevalence, a 0.6 transmission rate, a 0.5 recovery rate, and 7 days-incubation period. The population is fully connected, meaning agents can transmit the disease to any other agent:

```{r seir-conn}
model_seirconn <- ModelSEIRCONN(
  name                = "COVID-19",
  prevalence          = 0.01,
  n                   = 10000,
  contact_rate        = 10,
  incubation_days     = 7,
  transmission_rate   = 0.1,
  recovery_rate       = 1 / 7
) |> add_virus(
  virus(
    name = "COVID-19",
    prevalence = 0.01,
    as_proportion = TRUE,
    prob_infecting = 0.01,
    recovery_rate = 0.6,
    prob_death = 0.5,
    incubation = 7
))

set.seed(132)
run(model_seirconn, ndays = 100)
summary(model_seirconn)
```

Computing some key statistics

```{r seir-conn-figures}
plot(model_seirconn)

repnum <- get_reproductive_number(model_seirconn)

head(plot(repnum))

head(plot_generation_time(model_seirconn))
```


## SIR Logit

This model provides a more complex transmission and recovery pattern based
on agents' features. With it, we can reflect co-morbidities that could change
the probability of infection and recovery. Here, we simulate a population
including a dataset with two features: an intercept and a binary variable
`Female`. The probability of infection and recovery are functions of the intercept and
the `Female` variables. The following code simulates a population of 100,000 agents
in a small-world network. Each agent is connected to eight other agents. One
percent of the population has the virus, with an 80% chance of transmission.
Infected individuals recover at a 0.3 rate:

```{r logit-model}
# Simulating a population of 100,000 agents
set.seed(2223)
n <- 100000

# Agents' features
X <- cbind(
  Intercept = 1,
  Female    = sample.int(2, n, replace = TRUE) - 1
)

coef_infect  <- c(.1, -2, 2)
coef_recover <- rnorm(2)

# Creating the model
model_logit <- ModelSIRLogit(
  "covid2",
  data = X,
  coefs_infect      = coef_infect,
  coefs_recover     = coef_recover,
  coef_infect_cols  = 1L:ncol(X),
  coef_recover_cols = 1L:ncol(X),
  prob_infection = .8,
  recovery_rate = .3,
  prevalence = .01
)

# Adding a small-world population
agents_smallworld(model_logit, n, 8, FALSE, .01)

# Running the model
run(model_logit, 50)
plot(model_logit)

# Females are supposed to be more likely to become infected
rn <- get_reproductive_number(model_logit)

(table(
  X[, "Female"],
  (1:n %in% rn$source)
) |> prop.table())[, 2]

# Looking into the agents
get_agents(model_logit)
```

## Transmission network

This example shows how we can draw a transmission network from a simulation. The following code simulates a population of 500 agents in a small-world network. Each agent is connected to ten other agents. One percent of the population has the virus, with a 50% chance of transmission. Infected individuals recover at a 0.5 rate:

```{r transmission-net}
# Creating a SIR model
sir <- ModelSIR(
  name           = "COVID-19",
  prevalence     = .01,
  transmission_rate = .5,
  recovery       = .5
) |>
  # Adding a Small world population
  agents_smallworld(n = 500, k = 10, d = FALSE, p = .01) |>
  # Running the model for 50 days
  run(ndays = 50, seed = 1912)

# Transmission network
net <- get_transmissions(sir)

# Plotting
library(epiworldR)
library(netplot)
x <- igraph::graph_from_edgelist(
  as.matrix(net[, 2:3]) + 1
)

nplot(x, edge.curvature = 0, edge.color = "gray", skip.vertex = TRUE)
```

## Multiple simulations

`epiworldR` supports running multiple simulations using the `run_multiple` function. The following code simulates 50 SIR models with 1000 agents each. Each agent is connected to ten other agents. One percent of the population has the virus, with a 90% chance of transmission. Infected individuals recover at a 0.1 rate. The results are saved in a `data.frame`:

```{r multiple-example}
model_sir <- ModelSIRCONN(
  name = "COVID-19",
  prevalence = 0.01,
  n = 1000,
  contact_rate = 2,
  transmission_rate = 0.9, recovery_rate = 0.1
)

# Generating a saver
saver <- make_saver("total_hist", "reproductive")

# Running and printing
# Notice the use of nthread = 2 to run the simulations in parallel
run_multiple(model_sir, ndays = 100, nsims = 50, saver = saver, nthread = 2)

# Retrieving the results
ans <- run_multiple_get_results(model_sir)

head(ans$total_hist)
head(ans$reproductive)

plot(ans$reproductive)
```

# Tutorials

- The virtual INSNA Sunbelt 2023 session can be found here: https://github.com/UofUEpiBio/epiworldR-workshop/tree/sunbelt2023-virtual

- The in-person INSNA Sunbelt 2023 session can be found here:
https://github.com/UofUEpiBio/epiworldR-workshop/tree/sunbetl2023-inperson


# Citation

If you use `epiworldR` in your research, please cite it as follows:

```{r}
citation("epiworldR")
```

# Existing Alternatives

Several alternatives to `epiworldR` exist and provide researchers with a range of options, each with its own unique features and strengths, enabling the exploration and analysis of infectious disease dynamics through agent-based modeling. Below is a manually curated table of existing alternatives, including ABM [@ABM], abmR [@abmR], cystiSim [@cystiSim], villager [@villager], and RNetLogo [@RNetLogo].

| Package                                                                     | Multiple Viruses | Multiple Tools | Multiple Runs | Global Actions | Built-In Epi Models |  Dependencies                                                                                             | Activity                                                                                                               |
|:--------|:--------|:--------|:--------|:--------|---------|:--------|:--------|
| [**epiworldR**](https://cran.r-project.org/package=epiworldR)               | yes              | yes            | yes           | yes            | yes                 | [![status](https://tinyverse.netlify.com/badge/epiworldR)](https://CRAN.R-project.org/package=epiworldR) | [![Activity](https://img.shields.io/github/last-commit/UofUEpiBio/epiworldR)](https://github.com/UofUEpiBio/epiworldR) |
| [**ABM**](https://cran.r-project.org/package=ABM)                           | \-               | \-             | \-            | yes            | yes                 | [![status](https://tinyverse.netlify.com/badge/ABM)](https://CRAN.R-project.org/package=ABM)             | [![Activity](https://img.shields.io/github/last-commit/junlingm/ABM)](https://github.com/junlingm/ABM)                 |
| [**abmR**](https://cran.r-project.org/package=abmR)                         | \-               | \-             | yes           | \-             | \-                  | [![status](https://tinyverse.netlify.com/badge/abmR)](https://CRAN.R-project.org/package=abmR)           | [![Activity](https://img.shields.io/github/last-commit/bgoch5/abmR)](https://github.com/bgoch5/abmR)                   |
| [**cystiSim**](https://cran.r-project.org/package=cystiSim)                 | \-               | yes            | yes           | \-             | \-                  | [![status](https://tinyverse.netlify.com/badge/cystiSim)](https://CRAN.R-project.org/package=cystiSim)   | [![Activity](https://img.shields.io/github/last-commit/brechtdv/cystiSim)](https://github.com/brechtdv/cystiSim)       |
| [**villager**](https://cran.r-project.org/package=villager)                 | \-               | \-             | \-            | yes            | \-                  | [![status](https://tinyverse.netlify.com/badge/villager)](https://CRAN.R-project.org/package=villager)   | [![Activity](https://img.shields.io/github/last-commit/zizroc/villager)](https://github.com/zizroc/villager)           |
| [**RNetLogo**](https://cran.r-project.org/package=RNetLogo) | \-               | yes            | yes           | yes            | \-                  | [![status](https://tinyverse.netlify.com/badge/RNetLogo)](https://CRAN.R-project.org/package=RNetLogo)   | [![Activity](https://img.shields.io/github/last-commit/cran/RNetLogo)](https://github.com/cran/RNetLogo)               |


# Other ABM R packages

You may want to check out other R packages for agent-based modeling: [`ABM`](https://cran.r-project.org/package=ABM){target="_blank"},
[`abmR`](https://cran.r-project.org/package=abmR){target="_blank"},
[`cystiSim`](https://cran.r-project.org/package=cystiSim){target="_blank"},
[`villager`](https://cran.r-project.org/package=villager){target="_blank"}, and
[`RNetLogo`](https://cran.r-project.org/package=RNetLogo){target="_blank"}.

## Code of Conduct

The epiworldR project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
