---
title: "Social bubbles on a small-world network"
author:
  - George Vega Yon
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Social bubbles on a small-world network}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Social-bubble policies preserve contacts within small, fixed groups while
cutting down transmission between groups. How much of the outside contact is
cut is controlled by `transmission_factor`: `0` is a perfectly observed bubble
(no contact outside it at all), `1` leaves the population untouched, and values
in between represent partial compliance. This vignette compares an unmitigated
outbreak with two household-based policies created by `bubbles()`:

1. one household per bubble, and
2. up to three households per bubble.

Both policies operate from day 20 through day 59. In `bubbles()`, `end_day` is
exclusive, so `start_day = 20` and `end_day = 60` give that interval. The following diagram illustrates three potential bubbles:

![Three different bubbles that can be accomplished using the `bubbles()` function](img/bubble-small-diagram.svg){ width="80%"}

::: {.callout-important}
## Why this example uses SEIR

`ModelSIR()` has no exposed state and therefore cannot represent the requested
three-day incubation period. We use `ModelSEIR()`, the direct extension of the
SIR model, so that all the requested disease timings are represented. For a
strict SIR example, replace `ModelSEIR()` with `ModelSIR()` and omit
`incubation_days`.
:::

# Baseline outbreak

We simulate 10,000 agents for 100 days, beginning with 10 cases. The
small-world network has mean degree 40 and rewires 20% of its edges. With a
seven-day mean infectious period, the approximation

$$
R_0 \approx k \times \beta \times D
$$

gives a per-contact transmission probability of
$\beta = 2 / (40 \times 7)$. This is only an approximation on a clustered
network, which is why we describe the target $R_0$ as "close to 2."

```{r}
#| label: baseline-setup
library(epiworldR)

n_agents          <- 10000L
initial_cases      <- 10L
mean_degree        <- 40L
rewire_probability <- 0.20
incubation_days    <- 3
infectious_days    <- 7
target_r0          <- 2
ndays              <- 100L
nsims              <- 200L
simulation_seed    <- 2026L

transmission_rate <- target_r0 / (mean_degree * infectious_days)

baseline_model <- ModelSEIR(
  name              = "Example infection",
  prevalence        = initial_cases / n_agents,
  transmission_rate = transmission_rate,
  incubation_days   = incubation_days,
  recovery_rate     = 1 / infectious_days
)

agents_smallworld(
  baseline_model,
  n = n_agents,
  k = mean_degree,
  d = FALSE,
  p = rewire_probability
)
```

The network and disease parameters will be held fixed across policies. We also
use the same `simulation_seed` in each call to `run_multiple()`. The outbreak
size saver records the cumulative number infected, including the 10 initial
cases.

```{r}
#| label: helpers
run_scenario <- function(model, policy) {
  saver <- make_saver("outbreak_size")

  run_multiple(
    model,
    ndays    = ndays,
    nsims    = nsims,
    seed     = simulation_seed,
    saver    = saver,
    verbose  = FALSE,
    nthreads = 1L
  )

  outbreak <- run_multiple_get_results(
    model,
    nthreads = 1L
  )$outbreak_size

  final <- outbreak[outbreak$date == ndays, ]
  data.frame(
    policy        = policy,
    replicate     = final$sim_num,
    outbreak_size = final$outbreak_size
  )
}

summarize_outbreaks <- function(x) {
  pieces <- split(x$outbreak_size, x$policy)

  rows <- lapply(names(pieces), function(policy) {
    z <- pieces[[policy]]
    data.frame(
      policy        = policy,
      mean          = round(mean(z), 1),
      median        = unname(median(z)),
      lower_95_pct  = unname(quantile(z, 0.025)),
      upper_95_pct  = unname(quantile(z, 0.975)),
      mean_percent  = round(100 * mean(z) / n_agents, 1),
      row.names     = NULL
    )
  })

  do.call(rbind, rows)
}
```

# One household per bubble

First we generate household sizes from 2 through 6 with equal probability.
The only exception is at the end of the vector, where a draw that would leave
one unassigned agent is rejected. Household IDs are then assigned sequentially
in agent order.

This ordering is useful for a small-world network: before rewiring, nearby node
IDs are connected on the ring lattice. Sequential household members are
therefore likely to already share edges, even after 20% rewiring. This matters
because `bubbles()` restricts existing network ties; it does not create new
contacts between household members.

```{r}
#| label: households
set.seed(2026)
household_sizes <- sample_household_sizes(n_agents)
household_id <- rep(
  seq_along(household_sizes),
  times = household_sizes
)

stopifnot(
  length(household_id) == n_agents,
  all(household_sizes >= 2L & household_sizes <= 6L)
)

table(household_sizes)
```

For the first policy, each household is its own bubble (`group_size = 1`).
Contacts within a household are untouched -- those are what the policy
preserves. Contacts outside the household are reduced by 80% during the policy
window, meaning that 20% of the original per-contact probability remains
(`transmission_factor = 0.20`), which stands in for imperfect compliance rather
than a hermetic bubble.

```{r}
#| label: household-only-model
household_only_model <- clone_model(baseline_model)

bubbles(
  household_only_model,
  household_id        = household_id,
  flavor              = "household",
  group_size          = 1L,
  transmission_factor = 0.20,
  start_day           = 20L,
  end_day             = 60L
)
```

`bubbles()` does not keep the factor for itself: it stores it as a model
parameter, which the intervention reads on every exposure. Compliance can
therefore be inspected or changed -- between runs, or from a global event in the
middle of one -- without rebuilding the policy.

```{r}
#| label: transmission-factor-param
get_param(household_only_model, "Bubble transmission factor")
```

# Up to three households per bubble

The second policy changes only `group_size`. `bubbles()` joins connected
households into disjoint groups containing at most three households. The same
80% out-of-bubble reduction and day 20--60 policy window apply.

```{r}
#| label: three-household-model
three_household_model <- clone_model(baseline_model)

bubbles(
  three_household_model,
  household_id        = household_id,
  flavor              = "household",
  group_size          = 3L,
  transmission_factor = 0.20,
  start_day           = 20L,
  end_day             = 60L
)
```

# Compare outbreak sizes

Finally, we run 200 replicates for each scenario and report the final outbreak
size. Bubble models must use one thread because the intervention maintains
shared state; `bubbles()` marks them accordingly.

```{r}
#| label: run-scenarios
#| message: false
results <- rbind(
  run_scenario(baseline_model, "No bubbles"),
  run_scenario(household_only_model, "One household"),
  run_scenario(three_household_model, "Up to three households")
)

summarize_outbreaks(results)
```

The table reports the mean, median, central 95% simulation interval, and mean
percentage of the population ever infected. Comparing the two bubble policies
shows the trade-off directly: allowing households to combine preserves more
contacts, but it generally produces larger outbreaks than keeping each
household in its own bubble.
