---
title: "Using external network data: GeoPops synthetic populations"
author:
  - George Vega Yon
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Using external network data: GeoPops synthetic populations}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  fig.width  = 7,
  fig.height = 4.5,
  fig.align  = "center"
)
```

# Introduction

Most of the examples in epiworldR build their contact structure from a
random-graph generator---`agents_smallworld()`, `agents_sbm()`, or one of the
fully-connected `*CONN` models. That is convenient for prototyping, but many
applied questions start from a population that somebody else has already
constructed: a synthetic population, a mobility-derived contact network, a
survey-based sociocentric network, or a school roster.

This vignette shows the general recipe for that case, using data from the
[GeoPops](https://github.com/GeoPopsHub) project as the running example.[^data]
Nothing in the recipe is GeoPops-specific; the same four steps apply to any
population file plus edge list you may have:

[^data]: The GeoPops data used in this tutorial was provided by [Allisa Hamilton](https://github.com/alisahamilton1) and the GeoPops team. The extract is included in the `epiworldR` package for convenience, but the full Spartanburg County, SC population can be downloaded from [the GeoPops GitHub repository](https://github.com/GeoPopsHub/sc_spartanburg_measles/blob/1c741a362dc021b6df082d689f7f7c509f4c23b4/README.md).

1. Read the node table and the edge list(s).
2. Map the external node identifiers onto epiworld's agent identifiers, which
   are the contiguous integers `0, 1, ..., n - 1`.
3. Hand the re-indexed edge list to `agents_from_edgelist()`.
4. Attach a virus, run, and layer interventions on top with global events.

We then use the global events module (`?global-events`) to give the model a
weekly rhythm: transmission is dialed down on Saturday and Sunday and restored
on weekdays.

```{r}
#| label: load-pkg
library(epiworldR)
```

# The data

GeoPops generates geographically and demographically realistic synthetic
populations for U.S. Census geographies. A population comes as one *people*
table and one edge list per contact *layer*:

| File            | Contents                                                     |
|-----------------|--------------------------------------------------------------|
| `people_all.csv`| One row per person. `uid` is the unique person identifier.    |
| `net_h.csv`     | Household ties.                                              |
| `net_w.csv`     | Workplace ties.                                              |
| `net_s.csv`     | School ties.                                                 |
| `net_g.csv`     | Group-quarters ties.                                         |

Each edge file has columns `p1`, `p2`, and `edge_weight`, where `p1` and `p2`
are `uid`s from the people table. The full Spartanburg County, SC population
that this vignette is based on has 356,923 people and about 1.2 million ties.
That is far too large to ship inside an R package, so epiworldR includes a
much smaller extract: the largest census tract of the county, plus every
person directly tied to a resident of that tract, so that workplaces and
schools are not cut in half.

```{r}
#| label: data-path
geopops <- system.file("extdata", "geopops", package = "epiworldR")
list.files(geopops)
```

The files keep the GeoPops column layout, so the code below runs unchanged
against a full GeoPops download; only the file names change (drop the `.gz`).
`README.md` in that directory records exactly how the extract was cut, and
`data-raw/geopops-extract.R` in the package sources is the script that cuts it.

```{r}
#| label: read-people
people <- read.csv(file.path(geopops, "people_all.csv.gz"))

dim(people)
head(people[, c("uid", "hh_id", "tract", "age", "female", "working", "sch_grade")])
```

Two features of this table matter for what follows.

First, `uid` is *not* a row number here. In the full county the `uid`s happen
to run from 0 to 356,922 with no gaps, but our extract keeps only a subset of
them, so the identifiers are sparse:

```{r}
#| label: uid-gaps
range(people$uid)
nrow(people)
```

This is the normal situation with externally-supplied networks, and it is
exactly what step 2 of the recipe deals with.

Second, a block of people carries no demographic information at all:

```{r}
#| label: no-demog
table(demographics = ifelse(is.na(people$age), "missing", "present"))
```

These are people who work or attend school inside the study area but live
outside of it. GeoPops includes them so that workplace and school networks are
complete, but does not model their households. They are perfectly good agents
--- they just cannot be stratified by age or sex.

## The contact layers

We read the four edge lists and stack them into a single table, keeping track
of which layer each tie came from.

```{r}
#| label: read-edges
layers <- c(
  household = "net_h",
  workplace = "net_w",
  school    = "net_s",
  group     = "net_g"
)

edges <- do.call(rbind, lapply(names(layers), function(l) {
  e <- read.csv(file.path(geopops, paste0(layers[[l]], ".csv.gz")))
  data.frame(layer = l, p1 = e$p1, p2 = e$p2)
}))

table(edges$layer)
```

(For a full-county file, `data.table::fread()` or `vroom::vroom()` will read
these considerably faster than `read.csv()`.)

# From `uid`s to agent identifiers

epiworld numbers its agents `0, 1, ..., n - 1`, and `agents_from_edgelist()`
expects the `source` and `target` vectors to be in that numbering. So we build
a lookup from `uid` to position in the people table, and translate both
endpoints of every tie:

```{r}
#| label: remap
uid2id <- setNames(seq_along(people$uid) - 1L, people$uid)

edges$from <- uid2id[as.character(edges$p1)]
edges$to   <- uid2id[as.character(edges$p2)]

head(edges)
```

Because the lookup is keyed by the *row order* of `people`, agent `i` in the
model is row `i + 1` of `people` for the rest of this vignette. That is what
lets us go back and forth between simulation output and demographics.

A quick check that no tie points at somebody outside the people table:

```{r}
#| label: check-remap
any(is.na(edges$from) | is.na(edges$to))
```

## Collapsing the layers

For this model we treat the network as a single undirected graph: two people
are connected if they share a household, a workplace, a school, or a group
quarters. Since a pair can appear in more than one layer, we drop duplicated
pairs.

```{r}
#| label: dedupe
key <- paste(pmin(edges$from, edges$to), pmax(edges$from, edges$to))
net <- edges[!duplicated(key), ]

c(stacked = nrow(edges), unique_pairs = nrow(net))
```

We will need one more quantity later: the share of ties that are *not* work or
school ties, i.e. the ties that stay active on a weekend.

```{r}
#| label: home-share
home_ties  <- unique(key[edges$layer %in% c("household", "group")])
home_share <- length(home_ties) / nrow(net)
round(home_share, 3)
```

Roughly one tie in six survives the weekend; the rest of the network is
workplaces and schools.

# Loading the network into a model

Now the actual handoff. We create a `ModelSEIRD()` model --- exposed and
incubating, infectious, recovered, and dead, which is a reasonable skeleton
for COVID-19 --- and give it the GeoPops network. Note that `size` is the
number of *agents*, not the number of people with at least one tie: isolated
agents are legitimate members of the population and epiworld needs to know
about them.

```{r}
#| label: build-model
transmission_weekday <- 0.05

model <- ModelSEIRD(
  name              = "COVID-19",
  prevalence        = 0,
  transmission_rate = transmission_weekday,
  incubation_days   = 4,
  recovery_rate     = 1 / 7,
  death_rate        = 0.005
)

agents_from_edgelist(
  model,
  source   = net$from,
  target   = net$to,
  size     = nrow(people),
  directed = FALSE
)

model
```

The network is loaded, and we can read it back out with `get_network()`, which
returns the edge list as a two-column data frame of `from`/`to` agent ids:

```{r}
#| label: degree
loaded <- get_network(model)
nrow(loaded)

degree <- tabulate(c(loaded$from, loaded$to) + 1L, nbins = nrow(people))
summary(degree)
```

```{r}
#| label: degree-plot
#| fig-cap: "Degree distribution of the combined GeoPops contact network."
plot(
  table(degree),
  xlab = "Number of contacts",
  ylab = "Number of agents",
  main = "GeoPops contact network"
)
```

The shape is nothing like the tight, roughly Poisson degree distribution a
random graph would give at the same mean. Most of the mass sits at one to four
contacts---agents whose only ties are to their household---and then there is a
long shoulder stretching out past twenty, made up of people who also go to
work or to school. A few hundred agents have no ties at all. This kind of
structure is the whole reason for using an external population in the first
place.

## Seeding the outbreak from the people table

We set `prevalence = 0` above because we want to choose the index cases
ourselves rather than let epiworld scatter them at random. Here we start the
outbreak in twenty working adults, using the demographics to pick them:

```{r}
#| label: seed
set.seed(3312)
workers <- which(!is.na(people$working) & people$working == 1) - 1L
seeds   <- sample(workers, 20)

set_distribution_virus(
  get_virus(model, 0),
  distribute_virus_to_set(seeds)
)
```

Recall that `which(...) - 1L` converts R's 1-based row positions into
epiworld's 0-based agent identifiers. Fixing the seeds also makes the
comparison later in this vignette a fair one: both models start from exactly
the same twenty infections.

## A first run

```{r}
#| label: run-baseline
verbose_off(model)
run(model, ndays = 120, seed = 331)
summary(model)
```

```{r}
#| label: plot-baseline
#| fig-cap: "SEIRD dynamics on the GeoPops network, constant transmission rate."
plot(model)
```

# Adding a weekly rhythm with global events

The model above transmits at the same rate seven days a week. In reality, the
workplace and school layers---five ties out of six in this population---go
quiet on weekends.

epiworld's network is fixed once loaded, so we cannot delete those ties for
two days and add them back. What we can do is scale the per-contact
transmission probability. For small probabilities this is a good
approximation: the expected number of transmissions per infectious agent per
day is roughly (degree) x (transmission rate), so multiplying the rate by
`home_share` gives about the same expected transmissions as keeping only the
`home_share` fraction of ties at the full rate.

```{r}
#| label: weekend-rate
transmission_weekend <- transmission_weekday * home_share
c(weekday = transmission_weekday, weekend = round(transmission_weekend, 4))
```

## Timing

Global events run *after* the day's state updates. Concretely, on the step
where `today(model)` is `t`, transmission for day `t` has already happened
using whatever the parameter was set to previously; the event then decides
what day `t + 1` will use. So the event has to look one day ahead.

We take day 1 of the simulation to be a Monday, which makes the day of the
week `(day - 1) %% 7`, with 5 and 6 being Saturday and Sunday. Day 0 is the
baseline record that epiworld writes before the first step, so we exclude it:

```{r}
#| label: is-weekend
is_weekend <- function(day) (day >= 1L) & ((day - 1L) %% 7L) >= 5L

data.frame(
  day     = 0:8,
  weekend = is_weekend(0:8)
)
```

## The event

`globalevent_fun()` wraps an ordinary R function that receives the model and
is called at every step. Ours reads the clock and sets `"Transmission rate"`
for the following day:

```{r}
#| label: weekend-event
weekend_effect <- function(m) {
  set_param(
    m,
    "Transmission rate",
    if (is_weekend(today(m) + 1L)) transmission_weekend else transmission_weekday
  )
}
```

Day 1 is a weekday and no event has fired yet, so it uses the
`transmission_rate` we passed to `ModelSEIRD()`---which is why that argument
was set to the weekday value.

Now we rebuild the model, identically to before, and attach the event:

```{r}
#| label: build-weekend-model
model_weekend <- ModelSEIRD(
  name              = "COVID-19",
  prevalence        = 0,
  transmission_rate = transmission_weekday,
  incubation_days   = 4,
  recovery_rate     = 1 / 7,
  death_rate        = 0.005
)

agents_from_edgelist(
  model_weekend,
  source   = net$from,
  target   = net$to,
  size     = nrow(people),
  directed = FALSE
)

set_distribution_virus(
  get_virus(model_weekend, 0),
  distribute_virus_to_set(seeds)
)

add_globalevent(
  model_weekend,
  globalevent_fun(weekend_effect, "Weekend contact reduction")
)

verbose_off(model_weekend)
run(model_weekend, ndays = 120, seed = 331)
summary(model_weekend)
```

# Comparing the two runs

The clearest place to see the weekend effect is daily incidence---the number
of susceptible agents who become exposed each day---which we can pull out of
the transition matrix:

```{r}
#| label: incidence-fun
daily_incidence <- function(m) {
  tm <- get_hist_transition_matrix(m)
  tm <- tm[tm$state_from == "Susceptible" & tm$state_to == "Exposed", ]
  tm[order(tm$date), c("date", "counts")]
}

inc_baseline <- daily_incidence(model)
inc_weekend  <- daily_incidence(model_weekend)
```

```{r}
#| label: plot-incidence
#| fig-cap: "Daily new exposures. Shaded bands are weekends."
plot(
  inc_baseline$date, inc_baseline$counts,
  type = "l", col = "gray40", lwd = 2,
  xlab = "Day (day 1 = Monday)", ylab = "New exposures",
  main = "Daily incidence, with and without a weekend effect"
)

# Shading the weekends, one rectangle per Saturday-Sunday pair
saturdays <- with(inc_baseline, date[is_weekend(date) & (date - 1L) %% 7L == 5L])
for (d in saturdays) {
  rect(d - .5, -1e4, d + 1.5, 1e5, col = adjustcolor("steelblue", .12), border = NA)
}

lines(inc_weekend$date, inc_weekend$counts, col = "tomato", lwd = 2)
legend(
  "topright", bty = "n", lwd = 2, col = c("gray40", "tomato"),
  legend = c("Constant rate", "Weekday/weekend rate")
)
```

The red curve has a pronounced seven-day sawtooth: incidence collapses every
Saturday and Sunday and rebounds on Monday. Zooming in on two weeks around the
peak makes the pattern explicit:

```{r}
#| label: sawtooth-table
days <- 29:42
data.frame(
  day      = days,
  weekday  = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")[((days - 1) %% 7) + 1],
  constant = inc_baseline$counts[match(days, inc_baseline$date)],
  weekend  = inc_weekend$counts[match(days, inc_weekend$date)]
)
```

Beyond the weekly pattern, the intervention also flattens and delays the
outbreak overall:

```{r}
#| label: summary-compare
compare <- function(m, inc, label) {
  h        <- get_hist_total(m)
  fin      <- h[h$date == max(h$date), ]
  outbreak <- get_outbreak_size(m)
  outbreak <- outbreak[outbreak$date == max(outbreak$date), ]
  data.frame(
    model          = label,
    peak_incidence = max(inc$counts),
    peak_day       = inc$date[which.max(inc$counts)],
    ever_infected  = sum(outbreak$outbreak_size),
    deaths         = fin$counts[fin$state == "Deceased"]
  )
}

rbind(
  compare(model, inc_baseline, "Constant rate"),
  compare(model_weekend, inc_weekend, "Weekday/weekend rate")
)
```

## A thread-safe alternative

`globalevent_fun()` calls back into R, which is not thread safe, so
`run_multiple()` falls back to
`nthreads = 1` for any model that uses it. If you need many replicates in
parallel, you can get the same schedule out of `globalevent_set_params()`,
which is pure C++. Because the effect lands on the following day, the
"slow down" events go on Fridays and the "speed up" events on Sundays:

```{r}
#| label: setparams-alternative
ndays <- 120

model_weekend2 <- ModelSEIRD(
  name              = "COVID-19",
  prevalence        = 0,
  transmission_rate = transmission_weekday,
  incubation_days   = 4,
  recovery_rate     = 1 / 7,
  death_rate        = 0.005
)

agents_from_edgelist(
  model_weekend2,
  source = net$from, target = net$to,
  size = nrow(people), directed = FALSE
)

set_distribution_virus(
  get_virus(model_weekend2, 0),
  distribute_virus_to_set(seeds)
)

for (d in seq_len(ndays)) {

  new_rate <- if (!is_weekend(d) && is_weekend(d + 1L)) {
    transmission_weekend
  } else if (is_weekend(d) && !is_weekend(d + 1L)) {
    transmission_weekday
  } else {
    next
  }

  add_globalevent(
    model_weekend2,
    globalevent_set_params("Transmission rate", new_rate, day = d)
  )

}

verbose_off(model_weekend2)
run(model_weekend2, ndays = ndays, seed = 331)
```

The two formulations agree exactly:

```{r}
#| label: setparams-check
identical(
  get_hist_total(model_weekend),
  get_hist_total(model_weekend2)
)
```

# Scaling up to the full population

Everything above runs on the shipped extract, but epiworld is built for the
full thing. Loading the complete Spartanburg population---356,923 agents and
1,223,555 unique ties---takes well under a second, and a 100-day run takes
about the same. The code is identical; only the reading step changes:

```{r}
#| label: full-county
#| eval: false
library(data.table)

people <- fread("people_all.csv")

edges <- rbindlist(lapply(
  c("net_h", "net_w", "net_s", "net_g"),
  function(f) fread(paste0(f, ".csv"), select = c("p1", "p2"))
))

# In the full file the uids already run 0..n-1, so the remap is the identity;
# doing it anyway costs nothing and keeps the code portable.
uid2id <- setNames(seq_along(people$uid) - 1L, people$uid)
edges[, `:=`(from = uid2id[as.character(p1)], to = uid2id[as.character(p2)])]

net <- unique(edges[, .(a = pmin(from, to), b = pmax(from, to))])

model_full <- ModelSEIRD("COVID-19", 0, 0.05, 4, 1 / 7, 0.005)
agents_from_edgelist(
  model_full,
  source = net$a, target = net$b,
  size = nrow(people), directed = FALSE
)
```

# Caveats

A few things worth keeping in mind when you use this pattern on your own data.

- **The graph is static.** epiworld does not rewire the network between steps
  (other than the built-in degree-sequence rewiring), so layer-specific
  interventions such as school closures have to be expressed as parameter
  changes, as we did for the weekend, or by building a separate model per
  scenario.
- **Weights are ignored.** GeoPops ships an `edge_weight` column;
  `agents_from_edgelist()` takes an unweighted edge list. If your weights
  matter, one option is to encode them through agent- or virus-level
  functions rather than through the network itself.
- **Duplicated pairs.** Stacking layers can produce the same pair twice. We
  dropped duplicates above; leaving them in would have given those pairs two
  independent chances to transmit each day, which may or may not be what you
  want.
- **Boundary effects.** Our extract keeps one tract plus its immediate
  contacts, so agents on the boundary have artificially truncated
  neighborhoods. Results from the extract are illustrative, not an estimate
  for Spartanburg County.
