---
title: "Getting started with epiworldR"
author:
  - Derek Meyer
  - George Vega Yon
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with epiworldR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", out.width = "80%", fig.width = 7, fig.height = 5, 
  fig.align = "center"
)
```

# Introduction

epiworldR is an R package that provides a fast (C++ backend) and highly-
customizable framework for building network-based transmission/diffusion agent-
based models [ABM]. Some key features of epiworldR are the ability to construct 
multi-disease models (e.g., models of competing multi-pathogens/multi-rumor,) 
design mutating pathogens, architect population-level interventions, and build 
models with an arbitrary number of compartments/states (beyond SIR/SEIR.) 

# Example 1: Simulating an SIR model

## Setup and running the model

This example implements a social network with parameters listed within the 
ModelSIRCONN function. The disease name is specified (COVID-19), 100000 agents 
are initialized, disease prevalence of 0.001 is declared, each agent will 
contact two others (contact_rate), probability of transmitting the disease for 
any given agent is 0.3, and probability of recovery is set to $\frac{1}{3}$.
To create this model on epiworldR, simply use the ModelSIRCONN function. 
From here, the example will take you through the basic features of epiworldR.

```{r sirconn-setup}
library(epiworldR)
model_sir <- ModelSIRCONN(
  name              = "COVID-19",
  n                 = 100000, 
  prevalence        = 0.001, 
  contact_rate      = 2,
  prob_transmission = 0.3,
  prob_recovery     = 1/3
  )
```

Printing the model shows us some information. First, the name of the model, 
population size, number of entities (think of these as public spaces in which 
agents can make social contact with one another), the duration in days, number 
of variants, amount of time the last replicate took to run (last run elapsed t), 
and rewiring 
status (on or off). The next piece of information you will see is a list of the 
viruses used in the model. In this case, COVID-19 was the only disease used. 
Note that epiworldR has the capability to include more than one virus in a 
model. Tool(s) lists any tools that agents have to fight the virus. Examples of
this may include masking, vaccines, social distancing, etc. In this model, no 
tools are specified. Lastly, the model parameters are listed, which originate 
from the parameters specified in the model. 


```{r}
model_sir
```
  
To execute the model, use the run function with the SIR model object, number of 
simulation days, and an optional seed for reproducibility. Next, print out the 
results from the simulated model using model_sir. 
  
```{r}  
run(model_sir, ndays = 50, seed = 1912)
model_sir
```

There are two additional sections in the model summary after running the model 
object, the first being the distribution of the population at time 50. This 
section describes the flow of agents from each state (SIR) after 50 days. In the 
example, you'll see the number of agents in the susceptible state decreased
from 99900 to 26456, the number of agents in the infected state increased from
100 to 501, and the number of recovered agents increased to 73043 after 50 days. 
The counts for these states will of course, change based on model parameters or 
simulation run-time. The transmission probabilities section outputs a 3x3 matrix
that describes the probability of moving from one state to another. For example, 
in the susceptible row, each agent has a 0.97 probability of remaining in the 
susceptible state with a 0.03 probability of moving from the susceptible state 
to the infected state. Notice in all cases, there is a probability of 0 to skip 
states. In other words, it is impossible for an agent to jump from the 
susceptible state to the recovered state; that agent must pass through the 
infected state in order to then progress to the recovered state. The same logic
applies with moving backwards; an agent cannot become susceptible again after 
being infected. 
  
## Plot

## Extracting information

After running the epiworldR model, below is a list of all the functions that can 
be called using the epiworld model object. 

```{r showing-methods}
methods(class = "epiworld_model")
```

To demonstrate, start with the basic plot and get_hist_total functions. 

```{r}
plot(model_sir)
```

As evident from the above plot, the SIR model constructed from epiworldR 
displays the changes in susceptible, infected, and recovered case counts over 
time (days). Notice after a certain amount of time, the curves flatten. Below, 
a table representation of the above plot is printed, complete with each state 
within the SIR model, dates, and agent counts. 

```{r get-hist-total}
head(get_hist_total(model_sir))
```

An important statistic in epidemiological models is the reproductive number

```{r repnum}
repnum <- get_reproductive_number(model_sir)
head(repnum)
```

epiworldR has a method to automatically plot the reproductive number. This
function takes the average of values in the above table for each date and 
repeats until all dates have been accounted for. For example, on average, 
individuals who acquired the virus on the 10th day transmit the virus to roughly 
1.7 other individuals. 

```{r}
x <- plot(repnum, type="b")
```


## Adding more viruses/variants

epiworldR supports multi-variant models, the below code gives instructions on 
how to implement this. Using the "virus" function, give a name to the new 
virus/variant with its corresponding probability of infecting any given agent. 
In this example, prob_infecting is set to 1.0 as the flu is very contagious. 
In order to officially add this new virus to the model, use the "add_virus" 
function by calling the original epiworldR model object, the new virus, and 
the new virus' prevalence (which is set to 0.01 in this example).

```{r design-and-add}
# Building the virus
flu <- virus(name = "Flu", prob_infecting = 1.0)

# Adding the virus to the model
add_virus(model_sir, flu, .01)
```

After running the updated model with the new virus included for 50 days, the
output below describes the simulation. To confirm that the flu is included, 
notice the presence of "Flu" in the Virus(es) section of the output. All other
output is interpretable as specified in previous sections. 

```{r}
run(model_sir, ndays = 50, seed = 1912)
model_sir
```
Plotting the previous model (including the flu) yields the following. Notice 
the presence of two reproductive numbers plotted over time. Variant 0 refers 
to COVID-19 and variant 1 refers to the flu. 

```{r, fig.height=10}
repnum2 <- get_reproductive_number(model_sir)

op <- par(mfrow = c(2,1))
plot(model_sir)
plot(repnum2, type="b")
par(op)
```

## Tools

Now, the implementation of tools to combat any viruses and variants in the model 
will be demonstrated. First, for the sake of simplicity, remove the flu virus 
from the SIR model object (keep in  mind the index for the flu virus in the 
model object is 1). Next, provide parameters for the new tool using the "tool"
function. These parameters include the name of the tool, any reduction in 
probabilities for the SIR model parameters, and increased probability of 
recovery option. In order to add the tool to the SIR model, use the "add_tool"
function with the SIR model object, new tool, and prevalence of the tool. 
In this example, assume that 85% of the population will have received the 
vaccination. 

```{r}
# Removing the flu virus from the model
rm_virus(model_sir, 1)

vaccine <- tool(
  name = "Vaccine",
  susceptibility_reduction = .9,
  transmission_reduction = .5,
  recovery_enhancer = .5, 
  death_reduction = .9
)

add_tool(model_sir, vaccine, 0.85)
```

```{r curves-including-vaccine, fig.height=10}
repnum3 <- get_reproductive_number(model_sir)

op <- par(mfrow = c(2,1))
plot(model_sir)
plot(repnum3, type="b")
par(op)
```

