---
title: "Mixing models"
author:
  - George Vega Yon
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixing models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", out.width = "80%", fig.width = 7, fig.height = 5,
  fig.align = "center"
)
```

## Introduction

This vignette shows how to create a mixing model using the `epiworldR` package. Mixing models feature multiple populations. Each group, called `Entities`, has a subset of the agents. The agents can interact with each other within the same group and with agents from other groups. A contact matrix defines the interaction between agents.

$$
\left[%
\begin{array}{ccc}
0.9 & 0.05 & 0.05 \\
0.1 & 0.8 & 0.1 \\
0.1 & 0.2 & 0.7 \\
\end{array}%
\right]
$$


```{r model-setup} 
library(epiworldR)

e1 <- entity("Population 1")
e2 <- entity("Population 2")
e3 <- entity("Population 3")

# Row-stochastic matrix (rowsums 1)
cmatrix <- c(
  c(0.9, 0.05, 0.05),
  c(0.1, 0.8, 0.1),
  c(0.1, 0.2, 0.7)
) |> matrix(byrow = TRUE, nrow = 3)

N <- 1e4

flu_model <- ModelSEIRMixing(
  name              = "Flu",
  n                 = N,
  prevalence        = 10 / N,
  contact_rate      = 20,
  transmission_rate = 0.1,
  recovery_rate     = 1 / 7,
  incubation_days   = 7,
  contact_matrix    = cmatrix
)

flu_model |>
  add_entity_n(e1, N / 5) |>
  add_entity_n(e2, N / 5 * 2) |>
  add_entity_n(e3, N / 5 * 2)
```

```{r model-simulate}
set.seed(331)
run(flu_model, ndays = 100)
```

```{r model-summary}
summary(flu_model)
```

```{r viz}
plot_incidence(flu_model)
```

Investigating the history

```{r investigate, eval=FALSE}
library(data.table)

agents_entities <- lapply(get_entities(flu_model), \(e) {
  entity_get_agents(e)
}) |> rbindlist()

head(agents_entities)
```

```{r transmissions}
transmissions <- get_transmissions(flu_model) |>
  data.table()

transmissions <- subset(
  transmissions,
  select = c("date", "source")
)

transmissions <- merge(
  transmissions,
  agents_entities,
  by.x = "source", by.y = "agent"
)

# Aggregating by date x entity (counts)
transmissions <- transmissions[, .N, by = .(date, entity)]
```

```{r transmissions-plot}
setorder(transmissions, date, entity)

ran <- range(transmissions$N)
transmissions[entity == 0, plot(
  x = date, y = N, type = "l", col = "black", ylim = ran)]
transmissions[entity == 1, lines(x = date, y = N, col = "red")]
transmissions[entity == 2, lines(x = date, y = N, col = "blue")]

legend(
  "topright",
  legend = c("Population 1", "Population 2", "Population 3"),
  col = c("black", "red", "blue"),
  lty = 1
)
```